---
title: "Choosing cutoffs for assembly PEPMs"
output: html_notebook
---
## Set up environment
```{r}
library(tidyverse)
setwd("/Users/shane/Desktop/phosphonates/read-simulation/assembly_searches")
```
## Download the results to local dir
```{bash}
scp shogle@eofe4.mit.edu:/nobackup1b/users/shogle/projects/phosphonates/profile-search-vs-MARMICRODB/phn-synthesis/read_simulation/GEOTRACES-assemblies/search2_besthits.m8 GEOTRACES

scp shogle@eofe4.mit.edu:/nobackup1b/users/shogle/projects/phosphonates/profile-search-vs-MARMICRODB/phn-synthesis/read_simulation/TARA-OM-RGC/search2_besthits.m8 TARA-OM-RGC
```
## Load similarity files
```{r}
geotraces <- read_tsv("GEOTRACES/search2_besthits.m8", col_names=TRUE,  col_types="ccnnnnnnnn") 

tara <- read_tsv("TARA-OM-RGC/search2_besthits.m8", col_names=TRUE,  col_types="ccnnnnnnnn") 

MARMICRODB <- read_tsv("verifiedPEPM-all-v-all.m8", 
                       col_names=c("query", "target", "pident",
                                   "evalue", "bits", "qcov", "tcov",
                                   "qlen", "tlen", "alnlen"),
                       col_types="ccnnnnnnnn")

pepmsources <- read_tsv("pepm-summary-mod.tsv", col_names=c("id", "source", "seq"), col_types="ccc")
```
## PROCHLOROCOCCUS
```{r}
proMARMICRODB <- MARMICRODB %>% 
  filter(query %in% 
    (pepmsources %>% filter(source=="prochlorococcus") %>% pull(id))) %>% 
  group_by(query) %>%
  filter(query!=target) %>%
  arrange(desc(bits)) %>% 
  filter(bits > quantile(bits, .95)) %>%
  ungroup() %>%
  left_join(., pepmsources %>% rename(target=id)) %>%
  select(query, target, source, bits, pident, qcov, tcov)

proGEOTRACESids <- geotraces %>% 
  filter(target %in% (pepmsources %>% filter(source=="prochlorococcus") %>% pull(id))) %>% pull(query)

proGEOTRACES <- geotraces %>%
    filter(query %in% proGEOTRACESids)
```
## Write ids to collect sequences on engaging
```{r}
write_tsv(proGEOTRACES %>% 
  select(query) %>% 
  distinct(query) %>% 
  separate(query, c("sample", "scaffold", "gene"), "_", remove=FALSE) %>% 
  unite("scaffold", c("sample","scaffold"), sep="_", remove=FALSE) %>% 
  select(sample, scaffold, query, -gene), 
"pro_geotraces_scaffold_ids.tsv", col_names=FALSE)
```
## Tara Oceans (not sure if using tara yet... Stick with GEOTRACES for now)
```{r}
proTARAids <- tara %>% 
  filter(target %in% (pepmsources %>% filter(source=="prochlorococcus") %>% pull(id))) %>% pull(query)

proTARA <- tara %>%
  filter(query %in% proTARAids)
```
## PELAGIBACTERALES
```{r}
pelMARMICRODB <- MARMICRODB %>% 
  filter(query %in% 
           (pepmsources %>% filter(source=="pelagibacterales") %>% pull(id))) %>% 
  group_by(query) %>%
  filter(query!=target) %>%
  arrange(desc(bits)) %>% 
  filter(bits > quantile(bits, .95)) %>%
  ungroup() %>%
  left_join(., pepmsources %>% rename(target=id)) %>%
  select(query, target, source, bits, pident, qcov, tcov)

pelGEOTRACESids <- geotraces %>% 
  filter(target %in% (pepmsources %>% filter(source=="pelagibacterales") %>% pull(id))) %>% pull(query)

pelGEOTRACES <- geotraces %>%
  filter(query %in% pelGEOTRACESids) %>%
  left_join(., pepmsources %>% rename(target=id)) %>%
  select(query, target, source, bits, pident, qcov, tcov)
```
## Write ids to collect sequences on engaging
```{r}
write_tsv(pelGEOTRACES %>% 
            select(query) %>% 
            distinct(query) %>% 
            separate(query, c("sample", "scaffold", "gene"), "_", remove=FALSE) %>% 
            unite("scaffold", c("sample","scaffold"), sep="_", remove=FALSE) %>% 
            select(sample, scaffold, query, -gene), 
          "pel_geotraces_scaffold_ids.tsv", col_names=FALSE)
```
## Tara Oceans (not sure if using tara yet... Stick with GEOTRACES for now)
```{r}
pelTARAids <- tara %>% 
  filter(target %in% (pepmsources %>% filter(source=="pelagibacterales") %>% pull(id))) %>% pull(query)

pelTARA <- tara %>%
  filter(query %in% pelTARAids) %>%
  left_join(., pepmsources %>% rename(target=id)) %>%
  filter(qcov >= 0.85 & tcov >=0.85) %>%
  select(query, target, source, bits, pident, qcov, tcov)
```
## Send the id files to the cluster
```{bash}
scp pro_geotraces_scaffold_ids.tsv shogle@eofe4.mit.edu:/nobackup1b/users/shogle/projects/phosphonates/profile-search-vs-MARMICRODB/phn-synthesis/read_simulation/GEOTRACES-assemblies/initialsearch4pepm

scp pel_geotraces_scaffold_ids.tsv shogle@eofe4.mit.edu:/nobackup1b/users/shogle/projects/phosphonates/profile-search-vs-MARMICRODB/phn-synthesis/read_simulation/GEOTRACES-assemblies/initialsearch4pepm
```
## Grab the AA seqs and search against MARMICRODB
```{bash}
bioawk -t '{print $1,$2}' "pro_geotraces_scaffold_ids.tsv" | while read LIB ID; do seqkit fx2tab $ass/$LIB.genes.faa | grep $ID | seqkit tab2fx; done > pro_geotraces_scaffold.faa

mkdir tmp3
mmseqs createdb pro_geotraces_scaffold.faa PEPMDB/PEPMDB3 --dont-split-seq-by-len
mmseqs search PEPMDB/PEPMDB3 ${MARMICRODB} ALNDB/ALNDB3 tmp3 -a -s 7.5 --max-seqs 300 --threads 20
mmseqs convertalis PEPMDB/PEPMDB3 ${MARMICRODB} ALNDB/ALNDB3 search3.m8 --format-mode 0 --format-output 'query,target,pident,evalue,bits,qcov,tcov,qlen,tlen,alnlen'
```
## Download the results to local dir
```{bash}
scp shogle@eofe4.mit.edu:/nobackup1b/users/shogle/projects/phosphonates/profile-search-vs-MARMICRODB/phn-synthesis/read_simulation/GEOTRACES-assemblies/pro_search3.m8 .

scp shogle@eofe4.mit.edu:/nobackup1b/users/shogle/projects/phosphonates/profile-search-vs-MARMICRODB/phn-synthesis/read_simulation/GEOTRACES-assemblies/pel_search3.m8 .
```
## Load the results of searching the identified scaffolds vs MARMICRODB for Prochlorococcus
```{r}
pro_whole_scaffolds <- read_tsv("pro_search3.m8", 
                       col_names=c("query", "target", "pident",
                                   "evalue", "bits", "qcov", "tcov",
                                   "qlen", "tlen", "alnlen"),
                       col_types="ccnnnnnnnn") %>%
    group_by(query) %>%
    filter(bits==max(bits)) %>%
    left_join(., pepmsources %>% rename(target=id)) %>%
    separate(target, c("genome", "rest"), "__", remove=FALSE) %>%
    left_join(., genomeclass %>% rename(genome=sgenome)) %>%
    select(query, target, class, seq, pident, bits, evalue, qcov, tcov, qlen ,tlen, alnlen)
```

Valid Prochlorococcus PEPM (requires flanking genes also with best hit to Prochlorococcus genome)

Sample | PEPM sequence | Flanking genes | PEPM best hit to MARMICRODB 
------------- | ------------- | ------------- | -------------
S0191 | S0191_14299_1 | S0191_14299_1, S0191_14299_2 | AG-418-P13__01110_2182663
S0192 | S0192_16989_2 | S0192_16989_1, S0192_16989_2 | SB__01744_2182663
S0192 | S0192_24776_1 | S0192_24776_1 | AG-418-P13__01110_2182663
S0291 | S0291_13616_2 | S0291_13616_1, S0291_13616_2, S0291_13616_3 | AG-469-F22__00530_2182667
S0292 | S0292_27394_1 | S0292_27394_1, S0292_27394_2 | AG-469-F22__00530_2182667
S0318 | S0318_50549_2 | S0318_50549_1, S0318_50549_2 | AG-469-F22__00530_2182667
S0352 | S0352.2_25424_1 | S0352.2_25424_1 | AG-436-C13__00172_2182948
S0420 | S0420_7522_1 | S0420_7522_1, S0420_7522_2 | AG-469-F22__00530_2182667
S0422 | S0422_14583_2 | S0422_14583_1, S0422_14583_2 | AG-469-F22__00530_2182667
S0507 | S0507_35299_2 | S0507_35299_1, S0507_35299_2 | AG-347-I22__01066_2182663
S0540 | S0540_21234_1 | S0540_21234_1, S0540_21234_2 | AG-347-I22__01066_2182663
S0627 | S0627_13188_2 | S0627_13188_1, S0627_13188_2, S0627_13188_3 | AG-347-I22__01066_2182663

## Load the results of searching the identified scaffolds vs MARMICRODB for Prochlorococcus
```{r}
pel_whole_scaffolds <- read_tsv("pel_search3.m8", 
                       col_names=c("query", "target", "pident",
                                   "evalue", "bits", "qcov", "tcov",
                                   "qlen", "tlen", "alnlen"),
                       col_types="ccnnnnnnnn") %>%
    group_by(query) %>%
    filter(bits==max(bits)) %>%
    left_join(., pepmsources %>% rename(target=id)) %>%
    separate(target, c("genome", "rest"), "__", remove=FALSE) %>%
    left_join(., genomeclass %>% rename(genome=sgenome)) %>%
    separate(query, c("a", "b", "c"), "_", remove=FALSE) %>% 
    unite(scaffold, a, b, sep="_") %>% select(-c) %>%
    mutate(thresh=ifelse((seq=="pepm" & tcov<0.7), "bad", 
                         ifelse((seq=="pepm" & tcov>=0.9), "good", NA))) %>%
    arrange(query) %>% 
    group_by(scaffold) %>%
    fill(thresh, .direction = "down") %>%
    fill(thresh, .direction = "up") %>%
    select(scaffold, query, target, class, thresh, seq, pident, bits, evalue, qcov, tcov, qlen ,tlen, alnlen) %>% filter(thresh=="good")
```

Valid Pelagibacterales PEPM (requires flanking genes also with best hit to Pelagibacterales genome)

Sample | PEPM sequence | Flanking genes | PEPM best hit to MARMICRODB 
------------- | ------------- | ------------- | -------------
S0160 | S0160_15177_2 | S0160_15177_1, S0160_15177_2, S0160_15177_3, S0160_15177_4 | RS40__00847_2182711
S0351 | S0351_10706_1 | S0351_10706_1, S0351_10706_2 | AG-345-F21__00664_2183171
S0588 | S0588_11709_1 | S0588_11709_1, S0588_11709_2 | AG-325-E08__00885_2182690
S0622 | S0622_19920_2 | S0622_19920_1, S0622_19920_2, S0622_19920_3 | AG-430-I06__00923_2182722
S0624 | S0624_11743_2 | S0624_11743_1, S0624_11743_2 S0624_11743_3 | RS40__00847_2182711
S0624 | S0624_33737_1 | S0624_33737_1, S0624_33737_2 | RS40__00847_2182711
S0630 | S0630_45746_2 | S0630_45746_1, S0630_45746_2 | RS40__00847_2182711

