---
title: "Parametric modelling of Combined Bacteria/Archaea PepM abundance using top 5 environmental covariates identified using RFR"
output:   
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: png
    keep_html: yes
    html_preview: yes
---

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
library(corncob)
library(foreach)
library(doParallel)
library(RcppRoll)
library(ggforce)
```

```{r}
source(here("bin", "beta_binom_functions.R"))
```

# READ DATA
So the GEOTRACES data license is not GPL3 and you must request access and be approved before being allowed to download. So I am not including the GEOTRACES data here since it would be in violation of their license agreement. Unfortunately, this means that this is not truly reproducible as is. You can fill in the missing data for yourself by following the code available here - https://doi.org/10.5281/zenodo.3689249.

Preformatted dataset
```{r message=FALSE, warning=FALSE}
pepm <- readRDS(here::here("input", "pepm.global.final")) %>%
  group_by(sampleID, group) %>% 
  slice(1) %>%
  ungroup() %>%
  filter(sampleID != "S0468")
```

# VARIABLE SUBSETTING
## BACTERIA/ARCHAEA
```{r}
a <- left_join(filter(pepm, group=="bactarc") %>% select(sampleID, bactarc_pepm=RA),
               filter(pepm, group=="sar11") %>% select(sampleID, sar11_pepm=RA)) %>%
  mutate(sar11_pepm=ifelse(is.na(sar11_pepm), 1e-4, sar11_pepm)) %>%
  mutate(sar11_pepm=ifelse(sar11_pepm > 1, 0.99, sar11_pepm)) %>%
  mutate(bactarc_pepm=ifelse(is.na(bactarc_pepm), 1e-4, bactarc_pepm)) %>%
  mutate(bactarc_pepm=ifelse(bactarc_pepm > 1, 0.99, bactarc_pepm))

data2corncob <- pepm %>%
  filter(group=="bactarc") %>%
  drop_na() %>%
  mutate(RA=ifelse(RA > 1, 0.99, RA)) %>%
  mutate(RA=ifelse(is.na(RA), 1e-4, RA)) %>%
  mutate(W=ceiling(W),
         M=ceiling(M), 
         pel_IIb=pel_IIb*100,
         archaea=archaea*100,
         pel_IV=pel_IV*100,
         pro_LLIV=pro_LLIV*100) %>%
  left_join(., a)
```

# MODELS
```{r eval=FALSE, include=TRUE}
myfomula <- formula(cbind(W, M) ~ pel_IIb + sar11_pepm + archaea + pel_IV + pro_LLIV + depth) 

mymod <- bbdml(formula = myfomula,
               phi.formula = myfomula,
               link="logit",
               phi.link="logit",
               data = data2corncob)

summary(mymod)
```

### Bootstrap liklihood ratio tests for variables
- V1 = bactarc_pepm
- V2 = nitrateDarwin_dissolved_umol.kg
- V3 = pel_IV
- V4 = pel_IIb
- V5 = aluminum_dissolved_nmol.kg

Generate a null model omitting one of the terms
```{r eval=FALSE, include=TRUE}
null.1.f <- remove_terms(myfomula, "pel_IIb")
null.2.f <- remove_terms(myfomula, "sar11_pepm")
null.3.f <- remove_terms(myfomula, "archaea")
null.4.f <- remove_terms(myfomula, "pel_IV")
null.5.f <- remove_terms(myfomula, "pro_LLIV")
null.6.f <- remove_terms(myfomula, "depth")

myforms <- list(null.1.f, null.2.f, null.3.f, null.4.f, null.5.f, null.6.f)
      
names(myforms) <- c("V1", "V2", "V3","V4", "V5", "V6")

mynulls <- map(myforms, makenull, mymod, data2corncob)
```

```{r eval=FALSE, include=TRUE}
f.1.sim.stat <- PBLRTfun(mymod, mynulls$V1, 1000, bootLRT)
f.2.sim.stat <- PBLRTfun(mymod, mynulls$V2, 1000, bootLRT)
f.3.sim.stat <- PBLRTfun(mymod, mynulls$V3, 1000, bootLRT)
f.4.sim.stat <- PBLRTfun(mymod, mynulls$V4, 1000, bootLRT)
f.5.sim.stat <- PBLRTfun(mymod, mynulls$V5, 1000, bootLRT)
f.6.sim.stat <- PBLRTfun(mymod, mynulls$V6, 1000, bootLRT)
```

### Regular likelihood ratio tests
```{r eval=FALSE, include=TRUE}
f.1.lrt <- mypbLRT(mymod, mynulls$V1, f.1.sim.stat)
f.2.lrt <- mypbLRT(mymod, mynulls$V2, f.2.sim.stat)
f.3.lrt <- mypbLRT(mymod, mynulls$V3, f.3.sim.stat)
f.4.lrt <- mypbLRT(mymod, mynulls$V4, f.4.sim.stat)
f.5.lrt <- mypbLRT(mymod, mynulls$V5, f.5.sim.stat)
f.6.lrt <- mypbLRT(mymod, mynulls$V5, f.6.sim.stat)
```

### LRT results
results from bootstrapped LRT and regular LRT for the fully specified model
```{r eval=FALSE, include=TRUE}
lrt.res <- bind_rows(
  mutate(as_tibble(f.1.lrt), covariate="pel_IIb"),
  mutate(as_tibble(f.2.lrt), covariate="sar11_pepm"),
  mutate(as_tibble(f.3.lrt), covariate="archaea"),
  mutate(as_tibble(f.4.lrt), covariate="pel_IV"),
  mutate(as_tibble(f.5.lrt), covariate="pro_LLIV"),
  mutate(as_tibble(f.6.lrt), covariate="depth")
)

lrt.res

write_tsv(lrt.res, here::here("tables", "LRT_results.bactarc.tsv"))
```

# SIMULATIONS
Do the simulations in 100 replicate chunks. Don't know why but when you try 1000 for `sim_pa` it hangs. Breaking it up into chunks actually runs and finishes
```{r eval=FALSE, include=TRUE}
sims <- list()
  
for (i in 1:10) {
      sims[[i]] <- sim_pa(mymod, 100)
}

sims.drop <- lapply(sims, function(x) x[!is.na(x)])
```

```{r eval=FALSE, include=TRUE}
obs <- NULL
for (i in 1:10) {obs <- rbind(obs, sims[[i]][,1])}

mu <- NULL
for (i in 1:10) {mu <- rbind(mu, sims[[i]][,2])}

phi <- NULL
for (i in 1:10) {phi <- rbind(phi, sims[[i]][,3])}

summary.list <- sim_summary(mymod, list(obs, mu, phi))

obs2 <- summary.list[[1]]
mu2 <- summary.list[[2]] %>% rownames_to_column(var="variable")
phi2 <- summary.list[[3]] %>% rownames_to_column(var="variable")

save(obs2, mu2, phi2, file=here::here("data", "BB_sim.bactarc.Rdata"))
```