---
title: "Modelling prochlorococcus/SAR11 pepm abundance using environmental covariates"
output:   
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: png
    keep_html: yes
    html_preview: yes
---

# INTRO
useful tutorials:
http://www.rebeccabarter.com/blog/2020-03-25_machine_learning
https://juliasilge.com/blog/sf-trees-random-tuning/
https://www.tidymodels.org/start/case-study/
https://bradleyboehmke.github.io/HOML/random-forest.html

feature selection:
http://blog.datadive.net/selecting-good-features-part-iii-random-forests/
https://academic.oup.com/bib/article/20/2/492/4554516

SO questions on colinearity:
https://stats.stackexchange.com/questions/168622/why-is-multicollinearity-not-checked-in-modern-statistics-machine-learning
https://stats.stackexchange.com/questions/141619/wont-highly-correlated-variables-in-random-forest-distort-accuracy-and-feature

Evaluating performance
https://bookdown.org/max/FES/measuring-performance.html

```{r message=FALSE, warning=FALSE}
library(here)
library(tidyverse)

library(tidymodels)
library(recipeselectors)
library(vip)

#library(parallel)
#library(foreach)
#library(doParallel)

library(ranger)
library(Boruta)
library(care)

library(ggforce)
```

# READ DATA
Preformatted dataset
```{r message=FALSE, warning=FALSE}
pepm <- readRDS(here::here("input", "pepm.global.final")) %>%
  group_by(sampleID, group) %>% 
  slice(1) %>%
  ungroup() %>%
  filter(sampleID != "S0468")
```

# GENERAL FORMATTING
```{r eval=FALSE, include=TRUE}
a <- left_join(filter(pepm, group=="bactarc") %>% select(sampleID, bactarc_pepm=RA),
               filter(pepm, group=="sar11") %>% select(sampleID, sar11_pepm=RA)) %>%
  mutate(sar11_pepm=ifelse(is.na(sar11_pepm), 1e-4, sar11_pepm)) %>%
  mutate(sar11_pepm=ifelse(sar11_pepm > 1, 0.99, sar11_pepm)) %>%
  mutate(bactarc_pepm=ifelse(is.na(bactarc_pepm), 1e-4, bactarc_pepm)) %>%
  mutate(bactarc_pepm=ifelse(bactarc_pepm > 1, 0.99, bactarc_pepm))

PEPMreads2forest <- pepm %>%
  select_if(!str_detect(names(.), "imputed_")) %>%
  select_if(!str_detect(names(.), "pel_")) %>%
  select_if(!str_detect(names(.), "pro_LLVII")) %>% 
  select_if(!str_detect(names(.), "Long")) %>% 
  filter(group=="prochlorococcus") %>%
  select(-section, -lat, -lon, -group, -W, -M,
         -synDarwin_umolC.kg, -proDarwin_umolC.kg,
         -DOFe_darwin_clim, -POFe_darwin_clim) %>% 
  drop_na() %>%
  mutate(RA=ifelse(RA > 1, 0.99, RA)) %>%
  mutate(RA=ifelse(is.na(RA), 1e-4, RA)) %>%
  left_join(., a) %>%
  select(-sampleID,
         -bacteria,                       # redundant with archaea
         -pro_LLII_LLIII,                 # redundant with pro_LLIV
         -DOironDarwin_dissolved_nmol.kg, # redundant with DOPDarwin_dissolved_umol.kg
         -copper_dissolved_nmol.kg,       # nitrateDarwin_dissolved_umol.kg
         -ironDarwin_dissolved_nmol.kg)   # nitrateDarwin_dissolved_umol.kg
```

# DATA SPLITTING & RESAMPLING
```{r eval=FALSE, include=TRUE}
splits      <- initial_split(PEPMreads2forest, strata = RA, breaks=10)

pepm_other <- training(splits)
pepm_test  <- testing(splits)
```

# CROSS FOLD SET
```{r eval=FALSE, include=TRUE}
folds <- vfold_cv(pepm_other,
                  v = 10,
                  strata = RA, 
                  breaks = 10)
```

# RANDOM FOREST MODEL (RANGER) 
Perform with parallel processing
```{r eval=FALSE, include=TRUE}
cores <- parallel::detectCores()
cores
```

```{r eval=FALSE, include=TRUE}
rf_mod <- 
  rand_forest() %>%
  set_args(   mtry = tune(), 
              min_n = tune(), 
              trees = 1000) %>% 
  set_engine("ranger", 
             num.threads = cores,
             importance = "permutation", 
             splitrule = "beta",
             replace = FALSE) %>% 
  set_mode("regression")

rf_mod %>%    
  parameters()
```

# RECIPES
## CAR SCORE
This version contains the variable filter step using the CAR score algorithm from the `care` package. See here: http://www.strimmerlab.org/software/care/

The "care" package implements the regression approach of Zuber and Strimmer (2011). CAR scores measure the correlation between the response and the Mahalanobis-decorrelated predictors. The squared CAR score is a natural measure of variable importance and provides a canonical ordering of variables. This package provides functions for estimating CAR scores, for variable selection using CAR scores, and for estimating corresponding regression coefficients. Both shrinkage as well as empirical estimators are available.
```{r eval=FALSE, include=TRUE}
rf_recipe_carscore <- recipe(RA ~ ., data = pepm_other) %>%
  step_normalize(all_predictors(), -all_nominal()) %>%
  step_corr(all_predictors(), -all_nominal(), threshold = 0.9) %>%
  step_zv(all_predictors()) %>%
  step_select_carscore(all_predictors(), 
                       outcome = "RA", 
                       threshold = 0.5, # Retain features only within the top 50th percentile
                       top_p = tune())  # can be tuned
```

## BORUTA
```{r eval=FALSE, include=TRUE}
rf_recipe_boruta <- recipe(RA ~ ., data = pepm_other) %>%
  step_normalize(all_predictors(), -all_nominal()) %>%
  step_corr(all_predictors(), -all_nominal(), threshold = 0.9) %>%
  step_zv(all_predictors()) %>%
  step_select_boruta(all_predictors(), outcome = "RA", res="boruta_tune")
```

# WORKFLOW
## CARSCORE
```{r eval=FALSE, include=TRUE}
rf_workflow_carscore <- 
  workflow() %>% 
  add_recipe(rf_recipe_carscore) %>% 
  add_model(rf_mod)
```

### FEATURE SELECTION PARAMETERS
In case we are tuning top_p or threshold we need to update the tuned hyperparameters using upper bounds 
```{r eval=FALSE, include=TRUE}
p_info <- 
  rf_workflow_carscore %>% 
  parameters() %>% 
  update(threshold = threshold(c(0.5, 0.9))) %>%  # Pick an upper bound for the threshold
  update(top_p = top_p(c(1, 20))) %>%             # Pick an upper bound for number of variables 
  update(mtry = mtry(c(1, 10)))                   # Pick an upper bound for mtry 

ctrl <- control_grid(extract = identity,
                     verbose = TRUE)
```

## BORUTA
```{r eval=FALSE, include=TRUE}
rf_workflow_boruta <- 
  workflow() %>% 
  add_recipe(rf_recipe_boruta) %>% 
  add_model(rf_mod)
```

# TUNE HYPERPARAMETERS

## CARSCORE
```{r eval=FALSE, include=TRUE}
rf_res_carscore <-
  rf_mod %>% 
  tune_grid(
    rf_recipe,
    resamples = folds,
    grid = 20,
    param_info = p_info, # use only with the CARscore recipe
    control = ctrl,
    metrics = metric_set(rmse, rsq)
  )

saveRDS(rf_res_carscore, here::here("data", "rf_tune.carscore.pro"))
```

```{r eval=FALSE, include=TRUE}
rf_res_carscore %>% 
  show_best(metric = "rsq")
```

```{r eval=FALSE, include=TRUE}
rf_res_carscore %>% 
  show_best(metric = "rmse")
```

```{r eval=FALSE, include=TRUE}
autoplot(rf_res_carscore)
```

```{r eval=FALSE, include=TRUE}
rf_best_carscore <- 
  rf_res_carscore %>% 
  select_best(metric = "rsq")
rf_best_carscore
```

## BORUTA
```{r eval=FALSE, include=TRUE}
rf_res_boruta <-
  rf_mod %>% 
  tune_grid(
    rf_recipe,
    resamples = folds,
    grid = 20,
    control = ctrl,
    metrics = metric_set(rmse, rsq)
  )

saveRDS(rf_res_boruta, here::here("data", "rf_tune.boruta.pro"))
```

```{r eval=FALSE, include=TRUE}
rf_res_boruta %>% 
  show_best(metric = "rsq")
```

```{r eval=FALSE, include=TRUE}
rf_res_boruta %>% 
  show_best(metric = "rmse")
```

```{r eval=FALSE, include=TRUE}
autoplot(rf_res_boruta)
```

```{r eval=FALSE, include=TRUE}
rf_best_boruta <- 
  rf_res_boruta %>% 
  select_best(metric = "rsq")
rf_best_boruta
```

# FINAL FIT
After selecting our best model and hyperparameter values, our last step is to fit the final model on all the rows of data not originally held out for testing (both the training and the validation sets combined), and then evaluate the model performance one last time with the held-out test set. We’ll start by building our parsnip model object again from scratch. We take our best hyperparameter values from our random forest model.

## MODEL
Will run with Boruta
```{r eval=FALSE, include=TRUE}
final_rf_mod <- 
  rand_forest() %>%
  set_args(mtry = 9, 
              min_n = 4, 
              trees = 1000) %>% 
  set_engine("ranger", 
             num.threads = cores,
             importance = "permutation", 
             splitrule = "beta",
             replace = FALSE) %>% 
  set_mode("regression")
```

## RECIPE
```{r eval=FALSE, include=TRUE}
final_rf_recipe <- recipe(RA ~ ., data = pepm_other) %>%
  step_normalize(all_predictors(), -all_nominal()) %>%
  step_corr(all_predictors(), -all_nominal(), threshold = 0.9) %>%
  step_zv(all_predictors()) %>%
  step_select_boruta(all_predictors(), outcome = "RA", res="boruta_tune")
  #step_select_carscore(all_predictors(), -all_nominal(), outcome = "RA", top_p = 16)
```

## WORKFLOW
```{r eval=FALSE, include=TRUE}
final_rf_workflow <- 
  workflow() %>% 
  add_recipe(final_rf_recipe) %>% 
  add_model(final_rf_mod)
```

## EXECUTE
```{r eval=FALSE, include=TRUE}
final_rf_res <-
  final_rf_workflow %>% 
  last_fit(splits)

saveRDS(final_rf_res, here::here("data", "final_rf_res.pro"))
```

This fitted workflow contains everything, including our final metrics based on the test set. So, how did this model do on the test set? Was the validation set a good estimate of future performance?

```{r eval=FALSE, include=TRUE}
final_rf_res %>% 
  collect_metrics()
```

```{r eval=FALSE, include=TRUE}
final_rf_res %>% 
  pluck(".workflow", 1) %>%   
  pull_workflow_fit() %>% 
  vip(num_features = 59)
```

```{r eval=FALSE, include=TRUE}
final_rf_prep <- prep(final_rf_recipe)
boruta_obj <- final_rf_prep$steps[[4]]$res
```

```{r eval=FALSE, include=TRUE}
plot(boruta_obj)
```

```{r eval=FALSE, include=TRUE}
boruta_obj$ImpHistory
```
